On the effects of partying on next-day sports performance

### 2017-05-29 writing/draft.tex
Worked with Eric on first draft.

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. We are interested to see if 
teams exhibit a decline in performance the day following a game in a city 
with active nightlife. We exploit data on bookmaker spreads, the expected score 
differential between two teams after conditioning on observable performance. 
We expect a team to meet the spread half the time, since this is how bookmakers 
minimize risk on their end. We construct a model which attempts to 
estimate the causal effect of  visiting a "party-city" on subsequent day 
performance as measured by the spread. Since next-day
opponent is uncorrelated with exposure to treatment, we have identification in our variable of 
interest.

\section{Previous Research}

\section{Materials and Methods}
\subsection{Data}
\subsubsection{Lines data} We obtained data from \href{http://www.covers.com/sports/NBA/matchups?selectedDate=2011-1-01}{Covers.com}, which records the outcome of each game alongside the line set by a betting house.

@@

\subsubsection{Which cities (don't) have nightlife?}

We examine which cities have a notable nightlife, as measured by our proxy;
bottom and top five.

Los Angeles stands out, as does New York. Some runner ups include Chicago and Miami.

@@

\subsection{Possible confounders}
What if we pick up effects of a busy play schedule?
Aside from accounting for how much rest time a team is allowed in between games,
and travel distance between games, we're also interested to get a measure of
fatigue incurred on the court.
We account for this by bringing in minutes data, and counting the number of 
times the players run back and forth on the court. This is crudely measured
by the number of changes in posession. We back out this quantity.

\section{Results}

>>>>

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
Anecdotally, it has been reported that professional sports teams party on 
the road, sometimes to the detriment of the game.\cite{hoch}
J.R. Smith is infamous for citing his improvement in performance due to a
lack of nightlife when moving to Cleveland from New York.\cite{price,ley}
A few interviews with former professional athletes revealed that location
matters more than day of the week, i.e. parties can happen any day of the week,
but they are more likely to occur in select cities.

We are interested to see if 
teams exhibit a decline in performance the day following a game in a city 
with active nightlife. We exploit data on bookmaker spreads, the expected score 
differential between two teams after conditioning on observable performance. 
We expect a team to meet the spread half the time, since this is how bookmakers 
minimize risk on their end. We construct a model which attempts to 
estimate the causal effect of  visiting a "party-city" on subsequent day 
performance as measured by the spread. Since next-day
opponent is uncorrelated with exposure to treatment, we have identification in our variable of 
interest.

\section{Previous Research}
In 2014, Ezekowitz tried to answer the question we're interested in.\cite{ezekowitz} 
However, the study only
looks at performance in party cities, as opposed to subsequent day performance (i.e.
a hangover effect). Moreover, their sample size was much smaller. Finally, there was no
methodology used behind the selection of party cities. Between these factors, no
conclusion was reached with statistical significance. The data do weakly suggest
that teams are less likely to win on the road in party cities.

In 2015, Chase also looked at a statistical ranking of the biggest partiers in
the NBA.\cite{chase}

Studies on effects of alcohol on sports performance in a controlled setting.
Same goes for sleep deprivation and possibly narcotics. 
Both of those well correlated with active night life.

\section{Materials and Methods}
One of the hardest parts of this analysis is that partying is a latent variable.
In order to Determine when players partied, one might consider to look at Twitter data
for direct mentions of partying, either by athletes or witnesses. However, due to
intense public scrutiny that athletes face, there is a strong reluctance to post
such mentions on social media. There are notable exceptions, such
as Oddel Beckham Jr.'s trip to Miami in 2017 season.\cite{bleler}
The reports are few and far enough between,
and could be biased based on whether a particular city has nightclubs with VIP rooms,
that picking up statistical evidence would be a fruitless.

\paragraph{Two approaches: a discrete indicator and a continuous measure of nightlife}
Because famous athletes often enjoy a celebrity lifestyle, using conventional statistics
such as number of drinking establishments in each city does not work; athletes instead
tend to visit high end nightclubs. We again turned to player interviews for insight,
and found two major insights. We obtained a listing of cities which were anecdotally
mentioned as having active night-life. Between players, the rank ordering of these
was consistent to the extent that LA and NY were top
offenders. This motivates using a binary indicator for party in these two cities. 
We also found that athletes tend to party with other A-list celebrities,
particularly musicians. This motivated a continuous measure of nightlife, based on a
simple count of the number of musicians. We pursue both approaches.

Insights: major party cities are LA and NY, and athletes party is well correlated with musicians. This led to two separate definitions of party. Our most basic analysis was to consider performance 
the day after visiting LA or NY.
We also considered the number of musical establishments per city.

\paragraph{Logic for using spread data}
In order to assert a causal relation between cities with active nightlife 
and subsequent day sports performance, we need to control for salient
features such as the ability of players on either team in the match-up,
whether any players are injured, the momentum of success each team is carrying
in the season, etc. To that end, we utilize spread data. The benefit of
using spread data is that it controls for a lot, but the downside is that
we don't know exactly what it controls for.

For example, if book-keepers
are already aware of the effects of night-life on performance, it could
already be factored into the spread; if this were the case, then our variable
of interest would not appear to have significant explanatory power
on the margin, when also conditioning on spread, in spite of there being
a true effect. Alternatively,
if the house book-keeper systematically over-valued
teams which you are more likely to play after a party city,
we could see a spurious effect.
That said, the amount of money that is at stake and the effort put into creating these spreads 
suggests that they are reliable. So, if we do find an effect in our variable of interest,
we can be confident in our conclusion that there is a causal effect and not some
other systematic bias.

\subsection{Possible confounders}

\paragraph{Home team effect}
Home-team advantage is well-studied.\cite{jones07, jones08}  
Because teams play a lot of their games on the road,
most games after playing in a party city will be an away game.
This should be covered by spread data, but we also include
a home-team indicator to be sure.

\paragraph{Fatigue from previous game}
What if we pick up effects of a busy play schedule?
Aside from accounting for how much rest time a team is allowed in between games,
we're also interested to get a measure of
fatigue incurred on the court.
We account for this by bringing in minutes data, and counting the number of 
times the players run back and forth on the court. This is crudely measured
by the number of changes in posession. We back out this quantity.

\paragraph{Fatigue from Jet lag}
What about jet-lag? Prolonged air travel has in general been shown
to have a negative effect on sports performance.\cite{leeandgalvez}
Previous studies have shown that jet lag is a strong enough effect
to erase home field advantage in MLB.\cite{songetal}

\paragraph{The problem of an unknown travel schedule} In NBA, we have the difficulty of
not observing when teams travel. 
In MLB, each match-up is actually a series of games.

\subsection{Data}
\subsubsection{Lines data} We obtained data from \href{http://www.covers.com/sports/NBA/matchups?selectedDate=2011-1-01}{Covers.com}, which records the outcome of each game alongside the line set by a betting house.

@@

\subsubsection{Loctaion and Travel Distance Data} is obtained
by API.

@@

\paragraph{Discrete indicator}
If a team plays in Los Angeles or New York within 24 hours of their current game,
we indicate one, else zero.

\paragraph{Continuous measure}
If the last game was played at most 36 hours ago, use log average number
of musicians in last game location, else zero.

\subsection{Model} We use a logistic regression to estimate the likelihood of meeting
the spread as a function of whether the team partied as well as controlling for other confounders
listed above.

\section{Results}
\subsection{Using spread data}
\begin{center}
\begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]  \\[-1.8ex] & outcome == "W" \\  \hline \\[-1.8ex]   party & $-$0.052$^{**}$ (0.022) \\    lag.chg.pos & 0.0001 (0.002) \\    travel.dist & $-$0.00002 (0.00004) \\    ew & $-$0.035 (0.047) \\    nhours.lgame & 0.001 (0.001) \\    I(hour(date)) & 0.006 (0.012) \\    travel.dist:ew & 0.00004 (0.00005) \\    Constant & $-$0.144 (0.348) \\   \textit{N} & 9,979 \\  Log Likelihood & $-$6,912.426 \\  Akaike Inf. Crit. & 13,840.850 \\  \hline  \hline \\[-1.8ex]  \textit{Notes:} & \multicolumn{1}{r}{$^{***}$Significant at the 1 percent level.} \\   & \multicolumn{1}{r}{$^{**}$Significant at the 5 percent level.} \\   & \multicolumn{1}{r}{$^{*}$Significant at the 10 percent level.} \\  \end{tabular} 
\end{center}

\subsection{Using performance statistics}
We additionally looked at actual performance instead of relying on spread data. 
This allows us to avoid using the black-box of spread data, and also pin-point
more exactly what elements of the game are in fact effected by partying.
This motivates looking at number of points allowed, number of points scored.

We also considered rebounds, fouls, and even injuries. 
\begin{center}
    \begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]  \\[-1.8ex] & demeaned.reb.per.min \\  \hline \\[-1.8ex]   party & $-$0.001$^{*}$ (0.0005) \\    lag.chg.pos & 0.00003 (0.0001) \\    travel.dist & 0.00000$^{**}$ (0.00000) \\    nhours.lgame & 0.00002 (0.00002) \\    Constant & $-$0.003$^{**}$ (0.002) \\   \textit{N} & 44,190 \\  R$^{2}$ & 0.0003 \\  Adjusted R$^{2}$ & 0.0003 \\  Residual Std. Error & 0.105 (df = 44185) \\  F Statistic & 3.857$^{***}$ (df = 4; 44185) \\  \hline  \hline \\[-1.8ex]  \textit{Notes:} & \multicolumn{1}{r}{$^{***}$Significant at the 1 percent level.} \\   & \multicolumn{1}{r}{$^{**}$Significant at the 5 percent level.} \\   & \multicolumn{1}{r}{$^{*}$Significant at the 10 percent level.} \\  \end{tabular}  
\end{center}

\subsection{Robustness check}
We verify that this phenomena persists in the MLB. In MLB, money-lines
are used instead of point-spreads. So as opposed to predicting
the likelihood of meeting the spread, we predict the probability
a team would win against their opponent as both a function of the book-makers'
odds of winning and party. Additionally, we interacted weekend.
One of the major differences between NBA and MLB games is that MLB games
are played in series, so the same match-up repeated multiple times. This remedies
the problem of an unknown travel schedule we faced with NBA because we are sure 
teams stay overnight in the same location within a series.

\begin{center}
    \begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]  \\[-1.8ex] & I(team.score \textgreater  opponent.score) \\  \hline \\[-1.8ex]   party & $-$0.034$^{*}$ (0.019) \\    odds & 2.602$^{***}$ (0.162) \\    I(team == location) & 0.028 (0.033) \\    ndays.lgame & 0.012 (0.019) \\    travel.dist & 0.00000 (0.00003) \\    weekend & 0.042 (0.036) \\    Constant & $-$1.333$^{***}$ (0.080) \\   \textit{N} & 25,215 \\  Log Likelihood & $-$17,274.460 \\  Akaike Inf. Crit. & 34,562.920 \\  \hline  \hline \\[-1.8ex]  \textit{Notes:} & \multicolumn{1}{r}{$^{***}$Significant at the 1 percent level.} \\   & \multicolumn{1}{r}{$^{**}$Significant at the 5 percent level.} \\   & \multicolumn{1}{r}{$^{*}$Significant at the 10 percent level.} \\ \end{tabular}  
\end{center}

\subsubsection{Betting performance}
\begin{figure}[!h]
  \centering
  \includegraphics{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, The 
    circled point represents the worst out of pocket expense incurred during the season.
    We first train a model on historical data then use said model to predict on ``current year''
    data, ploting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet if our model instructs us of odds sufficiently different from what
    the betting house suggests. We account for the bookmaker's cut. E.g. in the early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have recouped all of
    our initial investment plus an additional \$11.5k.}
  \label{bettingperf}
\end{figure}

\subsection{Extending the research}
We can also validate where points are lost in MLB in a similar way.
It would be interesting to look at number of shots taken, and number of shots taken by stars.
It's possible there are prolonged effects of partying, and applying some sort of kernel
on the number of hours played since the last game could be interesting. It could also be
that different (types of) teams or players are more or less susceptible to the effects
of night-life; this could be demographic characteristics such as age or marital status.
